---
title: "The Impact of Credits, Revenue, and Audience Perception"
subtitle: "Box office Modeling"
output:
  html_document:
    toc: false 
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(httr)
library(jsonlite)
library(plotly)
library(dplyr)
library(tidyverse)
library(tidytext)
library(ggplot2)
library(knitr)
library(kableExtra)
library(car) 
library(MASS)
library(lubridate)
library(randomForest)
library(gridExtra)
library(broom)
library(tm)
library(textdata)
library(ggcorrplot)
library(patchwork)
library(htmltools)
library(DT)
```

```{r, include=FALSE}
knitr::opts_chunk$set(toc.depth = 1)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
movie_genres <- read.csv("data/movie_genres.csv")

reviews <- read.csv("data/reviews.csv")

movies <- read.csv("data/movies_full.csv")

movies <- movies %>%
  rename(tmdb_rating = vote_average)

movies <- movies %>%
  mutate(
    num_production_companies = str_count(production_companies, ", ") + 1,
    num_cast = str_count(cast, ", ") + 1,
    num_genre = str_count(genre_ids, ", ") + 1,
    num_production_countries = str_count(production_countries, ", ") + 1,
    first_genre = sapply(strsplit(as.character(genre_ids), ",\\s*"), `[`, 1)
  )

movies <- movies %>%
  mutate(first_genre = as.integer(first_genre)) %>%
  left_join(movie_genres, by = c("first_genre" = "id")) %>%
  mutate(first_genre = name) %>%
  dplyr::select(-name)

movies_cleaned_box <- movies %>%
  filter(!is.na(revenue) & !is.na(budget) & revenue > 100 & budget > 100)

movies_cleaned_box$log_revenue <- log(movies_cleaned_box$revenue)

reviews_clean <- reviews %>%
  filter(!is.na(content))

clean_text <- function(text) {
  text <- iconv(text, from = "UTF-8", to = "UTF-8", sub = "")
  text <- tolower(text)                
  text <- gsub("[^a-z'\\s]", " ", text)        
  text <- removeWords(text, stopwords("en"))   
  text <- stripWhitespace(text)             
  return(text)
}

reviews_clean <- reviews_clean %>%
  mutate(review = map_chr(content, clean_text)) |> 
  dplyr::select(review, movie_id)

reviews_clean <- reviews_clean %>%
  filter(str_trim(review) != "")



afinn_lex <- get_sentiments("afinn")

review_sentiment <- reviews_clean %>%
  mutate(review_id = row_number()) %>%
  unnest_tokens(word, review) %>%
  inner_join(afinn_lex, by = "word") %>%
  group_by(review_id, movie_id) %>%
  summarise(sentiment_score = mean(value), .groups = "drop")

movie_sentiment <- review_sentiment %>%
  group_by(movie_id) %>%
  summarise(avg_sentiment = mean(sentiment_score), .groups = "drop")

movie_sentiment_full <- movie_sentiment %>%
  inner_join(movies_cleaned_box, by = c("movie_id" = "id"))


model1 <- lm(log_revenue ~ budget + I(budget^2) + num_production_companies + num_cast +  num_production_countries + num_genre + tmdb_rating + imdb_rating + first_genre, data = movie_sentiment_full)

model2 <- lm(log_revenue ~ budget + I(budget^2) + num_production_companies + num_cast +  num_production_countries + num_genre + tmdb_rating + imdb_rating + first_genre + avg_sentiment, data = movie_sentiment_full)

set.seed(370)
train_index <- sample(1:nrow(movie_sentiment_full), 0.8 * nrow(movie_sentiment_full))
train_data <- movie_sentiment_full[train_index,]
test_data <- movie_sentiment_full[-train_index,]

rf_model1 <- randomForest(revenue ~ budget + num_production_companies + num_cast +  num_production_countries + num_genre + tmdb_rating + imdb_rating + first_genre,
                         data=train_data, ntree=100, importance=TRUE)
rf_model2 <- randomForest(revenue ~ budget + num_production_companies + num_cast +  num_production_countries + num_genre + tmdb_rating + imdb_rating + first_genre + avg_sentiment,
                         data=train_data, ntree=100, importance=TRUE)

custom_stopwords <- c("s", "t", "m", "ll", "ve", "re", "em", "d", "movie", "film", "films", "one", "two", "even", "just", "well", "also", "good", "will", "really", "sci", "fi", "can", "much", "oz", "ian")

reviews_genre <- reviews_clean %>%
  inner_join(movies_cleaned_box, by = c("movie_id" = "id"))
  
word_counts <- reviews_genre %>%
  unnest_tokens(word, review) %>%
  filter(!word %in% custom_stopwords) %>%
  count(first_genre, word, sort = TRUE)

tfidf_words <- word_counts %>%
  filter(n >= 5) |> 
  bind_tf_idf(word, first_genre, n) %>%
  arrange(desc(tf_idf))

```

# {.tabset}

## Regression

### Coefficient Table

This table presents the results of a multiple linear regression predicting log box office revenue based on key movie features

```{r echo=FALSE, message=FALSE, warning=FALSE}
model <- lm(log_revenue ~ budget + I(budget^2) + num_production_companies + num_cast +  num_production_countries + num_genre + tmdb_rating + imdb_rating + first_genre, data = movies_cleaned_box)

var_labels <- c(
  "(Intercept)" = "Intercept",
  "budget" = "Budget",
  "I(budget^2)" = "Budget Squared",
  "num_production_companies" = "Number of Production Companies",
  "num_cast" = "Number of Cast Members",
  "num_production_countries" = "Number of Production Countries",
  "num_genre" = "Number of Genres",
  "tmdb_rating" = "TMDB Rating",
  "imdb_rating" = "IMDB Rating",
  "first_genreAdventure" = "Genre: Adventure",
  "first_genreAnimation" = "Genre: Animation",
  "first_genreComedy" = "Genre: Comedy",
  "first_genreCrime" = "Genre: Crime",
  "first_genreDrama" = "Genre: Drama",
  "first_genreFamily" = "Genre: Family",
  "first_genreFantasy" = "Genre: Fantasy",
  "first_genreHistory" = "Genre: History",
  "first_genreHorror" = "Genre: Horror",
  "first_genreMusic" = "Genre: Music",
  "first_genreMystery" = "Genre: Mystery",
  "first_genreRomance" = "Genre: Romance",
  "first_genreScience Fiction" = "Genre: Science Fiction",
  "first_genreThriller" = "Genre: Thriller",
  "first_genreWar" = "Genre: War",
  "first_genreWestern" = "Genre: Western"
)

reg_table <- broom::tidy(model) %>%
  dplyr::select(term, estimate, std.error, statistic, p.value) %>%
  mutate(
    term = ifelse(term %in% names(var_labels), var_labels[term], term),
    significance = ifelse(p.value < 0.001, "very significant", 
                   ifelse(p.value < 0.01, "significant", "not significant")),
    estimate = formatC(estimate, format = "e", digits = 2),
    std.error = formatC(std.error, format = "e", digits = 2),
    statistic = round(statistic, 3),
    p.value = round(p.value, 3)
  )

datatable(
  reg_table,
  options = list(
    pageLength = 10,
    autoWidth = TRUE,
    scrollX = TRUE,
    columnDefs = list(
      list(width = '200px', targets = 0),  # term column wider
      list(width = '100px', targets = 1:4), # estimate, std.error, statistic, p.value
      list(width = '150px', targets = 5)    # significance
    )
  ),
  rownames = FALSE,
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: center; font-weight: bold;',
    'Table: Regression Results'
  )
)
```

### Insights

- Budget is the strongest and most significant predictor of revenue. Both budget and budget squared terms are highly significant, confirming a nonlinear relationship: increasing budgets boost revenue but with diminishing returns.

- Audience ratings from IMDB and TMDB show a positive, significant association with revenue, suggesting that higher-rated movies tend to perform better at the box office.

- Production companies (number involved) also have a notable effect, highlighting the role of production scale and marketing networks.

- Genre effects are mixed. While some genres (e.g., Action, Adventure) show differences compared to the baseline, most genre variables are not statistically significant in this model — possibly due to imbalances in genre representation.





## Random Forest

### Variable Importance

This plot compares feature importance from the random forest model.

```{r echo=FALSE, message=FALSE, warning=FALSE}
set.seed(370)
train_index2 <- sample(1:nrow(movies_cleaned_box), 0.8 * nrow(movies_cleaned_box))
train_data2 <- movies_cleaned_box[train_index2,]
test_data2 <- movies_cleaned_box[-train_index2,]

rf_model <- randomForest(revenue ~ budget + num_production_companies + num_cast +  num_production_countries + num_genre + tmdb_rating + imdb_rating + first_genre,
                         data=train_data2, ntree=100, importance=TRUE)

importance_df <- as.data.frame(importance(rf_model)) %>% 
  tibble::rownames_to_column(var = "Feature") %>%
  arrange(desc(`%IncMSE`))

var_labels <- c(
  num_cast = "Number of Casts",
  num_genre = "Number of Genres",
  num_production_companies = "Production Companies",
  num_production_countries = "Production Countries",
  budget = "Budget",
  revenue = "Revenue",
  imdb_rating = "IMDB Rating",
  imdb_votes = "IMDB Votes",
  tmdb_rating = "TMDB Rating",
  vote_count = "TMDB Vote Count",
  first_genre = "Main Genre"
)

importance_plot <- importance_df %>%
  mutate(
    Feature = ifelse(Feature %in% names(var_labels), var_labels[Feature], Feature)
  )

importance_plot$IncNodePurity <- as.numeric(importance_plot$IncNodePurity)

# === Plot for %IncMSE ===
plot_mse <- plot_ly(
  data = importance_plot,
  x = ~`%IncMSE`,
  y = ~reorder(Feature, `%IncMSE`),
  type = "bar",
  orientation = "h",
  marker = list(color = "steelblue"),
  text = ~paste("Feature:", Feature, "<br>%IncMSE:", round(`%IncMSE`, 2)),
  textposition = "none", 
  hoverinfo = "text",
  showlegend = FALSE
)

# === Plot for IncNodePurity ===
plot_purity <- plot_ly(
  data = importance_plot,
  x = ~IncNodePurity,
  y = ~reorder(Feature, IncNodePurity),
  type = "bar",
  orientation = "h",
  marker = list(color = "darkorange"),
  text = ~paste("Feature:", Feature, "<br>IncNodePurity:", round(IncNodePurity, 2)),
  textposition = "none", 
  hoverinfo = "text",
  showlegend = FALSE
)

subplot(plot_mse, plot_purity, nrows = 1, shareY = FALSE, titleX = TRUE, titleY = FALSE) %>%
  layout(
    title = list(
      text = "Random Forest Feature Importance Comparison",
      x = 0.5,
      xanchor = "center",
      font = list(size = 20)
    ),
    xaxis = list(
      title = list(text = "% Increase in MSE", standoff = 30),
      domain = c(0, 0.45)
    ),
    xaxis2 = list(
      title = list(text = "Increase in Node Purity", standoff = 30),
      domain = c(0.55, 1)
    ),
    yaxis = list(title = "", anchor = "x"),
    yaxis2 = list(title = "", anchor = "x2"),
    margin = list(t = 60),
    width = 800,
    height = 450 
  )

```

<div style="margin-top: 1em; font-size: 0.95em; color: #555;">
<strong>Figure 1:</strong> Random Forest Feature Importance Comparison based on two metrics: Percentage Increase in MSE and Increase in Node Purity.
</div>

### Insights

- Budget is by far the most important feature under both criteria, further confirming its dominant role in predicting movie success. But it has diminishing return.

- Audience ratings (IMDB and TMDB) and main genre also contribute meaningfully, although to a lesser extent.

- Cast size and number of production companies have moderate importance, suggesting that larger casts and production collaborations play supporting roles.

- Production countries and number of genres have minimal predictive value compared to other features.




## Model Comparison

### Box office Model Comparison: With vs Without Sentiment Score

This interactive plot compares the predictive performance of two modeling approaches—**Linear Regression** and **Random Forest**—in estimating movie box office revenue.

A toggle button above the plots allows you to switch between models that include **audience sentiment scores** as a feature and those that do not.

```{r sentiment_model_comparison, echo=FALSE, message=FALSE, warning=FALSE}
max_val <- max(movie_sentiment_full$revenue, na.rm = TRUE)
baseline_line <- list(
  type = "line",
  x0 = 0, y0 = 0,
  x1 = max_val, y1 = max_val,
  line = list(color = "gray", dash = "dash")
)

build_trace <- function(df, label, color, visible = TRUE) {
  list(
    x = df$actual,
    y = df$fitted,
    type = "scatter",
    mode = "markers",
    name = label,
    text = paste("Actual:", round(df$actual), "<br>Fitted:", round(df$fitted)),
    hoverinfo = "text",
    marker = list(color = color),
    visible = visible
  )
}

lm_with <- movie_sentiment_full %>% mutate(actual = revenue, fitted = exp(model2$fitted.values))
lm_without <- movie_sentiment_full %>% mutate(actual = revenue, fitted = exp(model1$fitted.values))
rf_train_with <- train_data %>% mutate(actual = revenue, fitted = predict(rf_model2, newdata = train_data))
rf_train_without <- train_data %>% mutate(actual = revenue, fitted = predict(rf_model1, newdata = train_data))
rf_test_with <- test_data %>% mutate(actual = revenue, fitted = predict(rf_model2, newdata = test_data))
rf_test_without <- test_data %>% mutate(actual = revenue, fitted = predict(rf_model1, newdata = test_data))

lm_w  <- build_trace(lm_with, "Linear Regression", "#1f77b4")
lm_wo <- build_trace(lm_without, "Linear Regression", "#1f77b4", FALSE)

rf_train_w  <- build_trace(rf_train_with, "Random Forest Train", "#2ca02c")
rf_train_wo <- build_trace(rf_train_without, "Random Forest Train", "#2ca02c", FALSE)

rf_test_w  <- build_trace(rf_test_with, "Random Forest Test", "#d62728")
rf_test_wo <- build_trace(rf_test_without, "Random Forest Test", "#d62728", FALSE)

max_axis <- max(movie_sentiment_full$revenue, na.rm = TRUE)

lm_plot <- plot_ly() %>%
  add_trace(x = lm_w$x, y = lm_w$y, type = lm_w$type, mode = lm_w$mode,
            name = lm_w$name, text = lm_w$text, hoverinfo = lm_w$hoverinfo,
            marker = lm_w$marker, visible = lm_w$visible) %>%
  add_trace(x = lm_wo$x, y = lm_wo$y, type = lm_wo$type, mode = lm_wo$mode,
            name = lm_wo$name, text = lm_wo$text, hoverinfo = lm_wo$hoverinfo,
            marker = lm_wo$marker, visible = lm_wo$visible) %>%
  layout(title = "Linear Model", shapes = list(baseline_line),
         xaxis = list(title = "Actual Revenue", range = c(0, max_axis)),
         yaxis = list(title = "Fitted Revenue", range = c(0, max_axis)))

rf_train_plot <- plot_ly() %>%
  add_trace(x = rf_train_w$x, y = rf_train_w$y, type = rf_train_w$type, mode = rf_train_w$mode,
            name = rf_train_w$name, text = rf_train_w$text, hoverinfo = rf_train_w$hoverinfo,
            marker = rf_train_w$marker, visible = rf_train_w$visible) %>%
  add_trace(x = rf_train_wo$x, y = rf_train_wo$y, type = rf_train_wo$type, mode = rf_train_wo$mode,
            name = rf_train_wo$name, text = rf_train_wo$text, hoverinfo = rf_train_wo$hoverinfo,
            marker = rf_train_wo$marker, visible = rf_train_wo$visible) %>%
  layout(title = "Random Forest (Train)", shapes = list(baseline_line),
         xaxis = list(title = "Actual Revenue", range = c(0, max_axis)),
         yaxis = list(title = "Fitted Revenue", range = c(0, max_axis)))

rf_test_plot <- plot_ly() %>%
  add_trace(x = rf_test_w$x, y = rf_test_w$y, type = rf_test_w$type, mode = rf_test_w$mode,
            name = rf_test_w$name, text = rf_test_w$text, hoverinfo = rf_test_w$hoverinfo,
            marker = rf_test_w$marker, visible = rf_test_w$visible) %>%
  add_trace(x = rf_test_wo$x, y = rf_test_wo$y, type = rf_test_wo$type, mode = rf_test_wo$mode,
            name = rf_test_wo$name, text = rf_test_wo$text, hoverinfo = rf_test_wo$hoverinfo,
            marker = rf_test_wo$marker, visible = rf_test_wo$visible) %>%
  layout(title = "Random Forest (Test)", shapes = list(baseline_line),
         xaxis = list(title = "Actual Revenue", range = c(0, max_axis)),
         yaxis = list(title = "Predicted Revenue", range = c(0, max_axis)))

full_plot <- subplot(lm_plot, rf_train_plot, rf_test_plot,
                     nrows = 1, shareY = TRUE, titleX = TRUE, titleY = TRUE) %>%
  layout(
    title = list(text = "Two Models Performance: With vs Without Sentiment", y = 0.83),
    width = 900,
    height = 400,
    margin = list(t = 80),
    updatemenus = list(
      list(
        type = "buttons",
        direction = "left",
        x = 0.4, y = 1.4,
        showactive = TRUE,
        buttons = list(
          list(method = "restyle", args = list("visible", rep(c(TRUE, FALSE), 3)), label = "With Sentiment Score"),
          list(method = "restyle", args = list("visible", rep(c(FALSE, TRUE), 3)), label = "Without Sentiment Score")
        )
      )
    )
  )

full_plot
```

<div style="margin-top: 1em; font-size: 0.95em; color: #555;">
<strong>Figure 2:</strong> Each panel displays an interactive scatter plot with a dashed 45° reference line indicating perfect prediction. Points closer to the diagonal suggest better model performance. This visual comparison helps assess whether including sentiment information improves prediction quality.
</div>

### Insights

- Random Forest models consistently outperform linear regression, capturing complex relationships between features and revenue.

- Adding sentiment scores provides minimal improvement. The model without sentiment performs similarly to the one with sentiment, suggesting that sentiment does not add significant predictive power beyond existing features like budget, ratings, and genre.

- This highlights that while audience emotions are interesting qualitatively, they may not be strong standalone predictors of financial success in a structured model.
